Modeling site-specific amino-acid preferences deepens phylogenetic divergence estimates for viral proteins

### 2018-04-16 manuscript/manuscript.tex
tweaks to empirical scatter figure
made empirical scatter figure 2 not 3 lines

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{figures/empirical_trees.pdf}}
\caption{\label{fig:empirical_trees}
\textbf{Effect of site-specific amino-acid preferences and $\Gamma\omega$ rate variation on HA branch length estimation.} 
The branch lengths of the HA tree were optimized using the indicated ExpCM or GY94 model. 
The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logo plots for 15 sites in HA; the full set of experimentally measured amino-acid preferences are in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}. 
The ExpCMs use amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
Circle denotes the H1 clade and triangle denotes the H3 clade.
The root of each tree is placed where it would fall if the tree was midpoint rooted using the branch lengths inferred by \texttt{RAxML} using the GTRCAT model. 
}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\textwidth]{figures/empirical_scatter.pdf}}
\caption{\label{fig:empirical_scatter}
\textbf{Modeling site-specific amino-acid preferences using deep mutational scanning experiments does not extend branch lengths uniformly across the HA tree.}
Points denote the branch lengths between all pairs of tips on the tree. 
Blue and orange denote branches that lead to the H1 and H3 deep mutational scanning reference sequences respectively. 
Each column is a different site-specific amino-acid preference set. 
(A) The ExpCMs defined by the H1, H3, or H1+H3 average deep mutational scanning amino-acid preferences estimate longer branches from the reference sequence used in the respective deep mutational scanning experiment. 
(B) The addition of $\Gamma\omega$ rate variation leads to more uniform extension of branch length than the addition of site-specific amino-acid preferences.
}
\end{figure}

@@

However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a roughly uniform fashion across the tree, the branch lengthening from adding site-specific amino-acid preferences is not uniform across the tree (\ref{fig:empirical_scatter}) 
Instead, the increase in branch length is most pronounced on branches leading to the HA sequence that was used in the deep mutational scanning experiment that informed the ExpCM.
For instance, the ExpCM informed by the H1 data most dramatically lengthens branches near the H1 clade of the tree, while the ExpCM informed by the H3 data has the largest effect on branches near the H3 clade.
The ExpCM informed by the average of the H1 and H3 data has a more uniform effect across the tree, but still most strongly extends branches leading to either the H1 or H3 clade.
Therefore, \ref{fig:empirical_trees} and \ref{fig:empirical_scatter} show that ExpCMs estimate longer branches, but that the effect is shaped by the set of amino-acid preferences used to inform the model.

>>>>

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{figures/empirical_trees.pdf}}
\caption{\label{fig:empirical_trees}
\textbf{Effect of site-specific amino-acid preferences and $\Gamma\omega$ rate variation on HA branch length estimation.} 
The branch lengths of the HA tree were optimized using the indicated ExpCM or GY94 model. 
The amino-acid preferences defining the model (ExpCM) or implied by the model (GY94) are shown as logo plots for 15 sites in HA; the full set of experimentally measured amino-acid preferences are in \ref{suppfig:prefs_doud}, \ref{suppfig:prefs_lee}, and \ref{suppfig:prefs_average}. 
The ExpCMs use amino-acid preferences measured in deep mutational scanning of an H1 HA~\citep{doud2016accurate}, an H3 HA~\citep{lee2018deep}, or the average of the measurements for these two HAs.
Circle denotes the H1 clade and triangle denotes the H3 clade.
The root of each tree is placed where it would fall if the tree was midpoint rooted using the branch lengths inferred by \texttt{RAxML} using the GTRCAT model. 
The branch lengths are plotted for quantitative comparison in \ref{fig:empirical_scatter}. 
}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{figures/empirical_scatter.pdf}}
\caption{\label{fig:empirical_scatter}
\textbf{Modeling site-specific amino-acid preferences using deep mutational scanning experiments does not extend branch lengths uniformly across the HA tree.}
Points denote the branch lengths between all pairs of tips on the tree. 
Blue and orange denote branches that lead to the H1 and H3 deep mutational scanning reference sequences respectively. 
The site-specific amino-acid preferences defining the ExpCM's are labeled above each plot.
These data are the branch lengths in \ref{fig:empirical_trees} plotted quantitatively for comparison.
(A) The ExpCMs defined by the H1, H3, or H1+H3 average deep mutational scanning amino-acid preferences estimate longer branches from the reference sequence used in the respective deep mutational scanning experiment. 
(B) The addition of $\Gamma\omega$ rate variation leads to more uniform extension of branch length than the addition of site-specific amino-acid preferences.
}
\end{figure}

@@

However, while adding $\Gamma\omega$ rate variation increases the length of deep branches in a roughly uniform fashion across the tree, the branch lengthening from adding site-specific amino-acid preferences is not uniform across the tree (\ref{fig:empirical_scatter}). 
Instead, the increase in branch length is most pronounced on branches leading to the HA sequence that was used in the deep mutational scanning experiment that informed the ExpCM.
For instance, the ExpCM informed by the H1 data most dramatically lengthens branches near the H1 clade of the tree, while the ExpCM informed by the H3 data has the largest effect on branches near the H3 clade.
The ExpCM informed by the average of the H1 and H3 data has a more uniform effect across the tree, but still most strongly extends branches leading to either the H1 or H3 clade.
Therefore, \ref{fig:empirical_trees} and \ref{fig:empirical_scatter} show that ExpCMs estimate longer branches, but that the effect is shaped by the set of amino-acid preferences used to inform the model.
